---
title: "Income Convergence in CESEE since EU Accession"
subtitle: "Beta-convergence and a staggered event-study analysis"
author:
  - name: "wiiw Research (template)"
    affiliation: "The Vienna Institute for International Economic Studies (wiiw)"
date: last-modified
abstract: |
  We track real income convergence in the eleven Central, East and Southeast
  European (CESEE) members of the European Union since their accession waves of
  2004, 2007 and 2013. A cross-sectional beta-convergence regression documents
  catch-up relative to long-standing members, and a staggered Sun & Abraham
  event study estimates the trajectory of the convergence gap around accession.
format:
  # PDF output is disabled for the seminar demo because it requires a LaTeX
  # install (TinyTeX). Re-enable by uncommenting this block after running
  # `quarto install tinytex` on the demo machine.
  # pdf:
  #   documentclass: article
  #   papersize: a4
  #   geometry: margin=2.5cm
  #   number-sections: true
  #   colorlinks: true
  #   fig-pos: "H"
  html:
    toc: true
    number-sections: true
    embed-resources: true
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(targets)
library(fixest)          # for aggregate.fixest dispatch inside event_study_att()
source("R/config.R")
source("R/provenance.R")
source("R/models.R")     # event_study_att(), called by tab_main() for Table 1
source("R/tables.R")

est <- tar_read(estimates)
```

# Introduction

The economic case for European Union enlargement rests substantially on
convergence, the expectation that poorer entrants catch up to the incumbent
average through trade, capital deepening, institutional anchoring and access to
the single market [@Barro1992]. This paper quantifies that catch-up for the
`r est$n_cesee` CESEE members that acceded in the waves of 2004, 2007 and 2013,
using `r est$n_control` long-standing members as a never-treated comparison
group.

# Data

We draw real GDP per capita (chain-linked volumes, `r est$gdp_unit`) for the
years `r est$sample_start` to `r est$sample_end` from Eurostat table
`nama_10_pc`. The outcome for the event study is the *convergence gap*, the log
of a country's real GDP per capita minus the log of the EU aggregate, so that a
rising series denotes catch-up. Country and year fixed effects absorb the common
European trajectory.

# Empirical strategy

We report two complementary objects. First, a cross-sectional beta-convergence
regression relates each country's average annual growth of real GDP per capita
since `r CONV_BASE` to its initial log level. A negative slope is convergence.
Second, a staggered difference-in-differences event study identifies the effect
of EU accession on the convergence gap. Because the three accession waves imply
staggered treatment timing, we use the Sun & Abraham (2021) interaction-weighted
estimator [@SunAbraham2021], which is robust to the heterogeneity-across-cohorts
bias of the conventional two-way fixed-effects specification. Identification
comes from the research design, the timing of accession relative to
never-treated members, not from the estimator.

# Results {#sec-results}

Figure 1 shows the classic convergence pattern. Countries that began the period
poorer grew faster. The fitted beta slope is `r fmt_num(est$beta_coef)`
(standard error `r fmt_num(est$beta_se)`), and the CESEE members sit clearly in
the high-growth, low-initial-income quadrant.

The event study in Figure 2 traces the convergence gap around accession. Pre-
accession coefficients are flat, consistent with parallel pre-trends. After
accession the gap narrows steadily. By the fifth year the convergence gap had
closed by `r fmt_lp(est$accession_effect_5y)` log points relative to the
year before accession, and by the tenth year by `r fmt_lp(est$accession_effect_10y)`
log points. Averaged over the post-accession horizon, the accession effect on
the convergence gap is `r fmt_lp(est$accession_att)` log points (Table 1).

```{r}
#| label: fig-convergence
#| fig-cap: "Income convergence across EU member states."
#| fig-width: 6.5
#| fig-height: 4.2
# Print the stored ggplot so knitr renders and Quarto embeds it directly. This
# avoids the file-path mismatch under execute-dir: project (tar_read of the PNG
# path resolves against the project root, but Quarto embeds relative to this
# .qmd's folder), which left the image unresolved and unembedded.
tar_read(fig1)
```

```{r}
#| label: fig-eventstudy
#| fig-cap: "Effect of EU accession on the convergence gap (Sun & Abraham)."
#| fig-width: 6.5
#| fig-height: 4.2
tar_read(fig2)
```

```{r}
#| label: tbl-main
#| tbl-cap: "Convergence and the accession effect."
tab_main(tar_read(conv_fit), tar_read(es_fit))
```

# Discussion

The estimates reproduce the received wisdom that EU accession is associated with
measurable income catch-up. Natural extensions include a wider set of countries,
an alternative price basis, and the @CallawaySantAnna2021 estimator as a
robustness check on the event-study results.

# References {.unnumbered}

::: {#refs}
:::
